The Creation of a Programming Language

### 2022-10-23 paper.tex
continue writing paper

@@

% Include C code from file #1 from lines #2 to #3
\newcommand{\code}[3] {
    \ver{
    \lstset{basicstyle=\scriptsize,keywordstyle=\color{blue},stepnumber=5,
        stringstyle=\color[rgb]{.3,.5,.1},commentstyle=\color{gray},
        morecomment=[l][\color{magenta}]{\#},
        postbreak=\mbox{\textcolor{red}{$\hookrightarrow$}\space}}
    \lstinputlisting[language=C,numbers=left,breaklines=true,
        frame=single,title=#1,
        firstnumber=#2,firstline=#2,lastline=#3]{#1}
    }
}

@@

\section{Theory}
\subsection{Different Types of Interpreters}
In computing, there are several types of interpreters that behave differently.
Here are some examples and their use-cases:
\begin{itemize}
    \item \emph{Just in time Compiler.} JIT-compilers blur the line between
        classical interpreters and compilers. As the name suggests, code is
        being compiled at run time of the program. The JIT-compiler can optimize
        compilation of more-often used code and therefore adapt itself to
        behave somewhat close to optimally for a given use-case. Today, this 
        technique is often used instead of classical interpreters, as
        JIT-compilers come with only advantages. The most popular example of
        a JIT-compiler would be Google's V8 JavaScript engine used
        in NodeJS and any Chromium browsers.
    \item \emph{Bytecode Interpreter.} Such interpret the bytecode that was
        output by a bytecode compiler from a given piece of code. The bytecode
        is an intermediate representation of the program, used to speed up
        both compilation and interpretation. Such an intermediary bytecode is
        useful for ensuring platform independence or portability, as it will be
        the same across any machine, the only difference being the interpreter.
        The most popular example would be the Java Virtual Machine,
        a bytecode interpreter for java bytecode, also used by many other languages
        such as Kotlin, Scala or Groovy.
    \item \emph{Abstract Syntax Tree interpreter.} Source code can be
        transformed into an AST or parse tree by a parser. A simple example
        can be found under 
        \hyperref[simple interpreter]{'Implementation/The Art of Abstraction'}
        on page \pageref{simple interpreter}. Compared to bytecode interpreters
        there is a large time overhead when performing syntax related computation
        or visiting tree nodes recursively.
    \item \emph{Self interpreter.} These are interpreters that interpret the
        language they themselves are written in. This requires a program
        written in another language running an interpreter for the wanted
        language, in which runs another interpreted for and written in said
        language. Using a host language to initiate such system is called
        bootstrapping.

@@

The problem to solve in this paper is concisely to make a computer understand
human-written and -readable text, called source code, and follow the instructions
it proposes.
Most interpreters follow a plain scheme to go about this problem, but the
actual implementation of each of the steps can be heavily customized.
Here is the problem, broken down into three steps as commonly done in other
popular programming languages:
\begin{enumerate}\pagelabel{simple interpreter}
    \item To create meaningful groups of characters from the input string. This
        process is called 'tokenization' and is done with the component called
        tokenizer or lexer. As an example, the string 'aa + bb' could result
        in three tokens, namely 'aa', '+' and 'bb'. In this case, space characters
        are ignored. 
        This is only an example and the grouping of characters could be done 
        in any thinkable, logical way.
    \item To order the tokens into parse-tree, also called syntax-tree, as
        specified by rules. These rules are called a grammar.
        Creating an order is important, as some calculations depend on others, e.g. multiplication
        is always done before addition. An IT data structure that can accommodate
        such requirements is the binary tree. A binary tree consists of tree-nodes
        that each have a maximum of two other tree nodes as their children, and
        a tree node whose child they are, the parent node.
    \item To interpret the parse tree, traversing through it and following the
        instructions each of the tree-nodes represent. The outcome
        will be the result of the computation the program describes.
\end{enumerate}

@@

Here is a file tree overview of the project which will be referenced
several times in the following sections of this paper. In the depicted file tree
only source files are shown. No build files, executable files or files that
are of no further importance regarding \name are depicted.
\newcommand{\hr}[2]{\hyperref[#2]{#1}}
\ver{\begin{flushleft}
    \scriptsize\setlength{\DTbaselineskip}{7pt}
    \dirtree{%
        .1 project root.
        .2 \hr{.gitignore}{.gitignore}.
        .2 \hr{README.md}{README.md}.
        .2 \hr{Makefile}{Makefile}.
        .2 \hr{mapop.gperf}{mapop.gperf}.
        .2 pseudocode.
            .3 \hr{parser.c}{pseudocode/parser.c}.
        .2 examples.
            .3 \hr{mandelbrot.nts}{examples/mandelbrot.nts}.
            .3 \hr{fibonacci.nts}{examples/fibonacci.nts}.
            .3 \hr{exec.nts}{examples/exec.nts}.
        .2 src.
            .3 \hr{main.c}{src/main.c}.
            .3 interpreter.
                .4 \hr{bytecode.c}{src/interpreter/bytecode.c}.
                .4 \hr{bytecode.h}{src/interpreter/bytecode.h}.
                .4 error.
                    .5 \hr{error.c}{src/interpreter/error/error.c}.
                    .5 \hr{error.h}{src/interpreter/error/error.h}.
                .4 \hr{interpreter.c}{src/interpreter/interpreter.c}.
                .4 \hr{interpreter.h}{src/interpreter/interpreter.h}.
                .4 libraries.
                    .5 \hr{libraries.c}{src/interpreter/libraries/libraries.c}.
                    .5 \hr{libraries.h}{src/interpreter/libraries/libraries.h}.
                    .5 stdlib.
                        .6 \hr{stdlib.c}{src/interpreter/libraries/stdlib/stdlib.c}.
                        .6 \hr{stdlib.h}{src/interpreter/libraries/stdlib/stdlib.h}.
                .4 \hr{localizer.c}{src/interpreter/localizer.c}.
                .4 \hr{localizer.h}{src/interpreter/localizer.h}.
                .4 mappings.
                    .5 \hr{operations.c}{src/interpreter/mappings/operations.c}.
                    .5 \hr{operations.h}{src/interpreter/mappings/operations.h}.
                    .5 \hr{mapop.c}{src/interpreter/mappings/mapop.c}.
                    .5 \hr{mapop.h}{src/interpreter/mappings/mapop.h}.
                .4 memory.
                    .5 \hr{hashtable.c}{src/interpreter/memory/hashtable.c}.
                    .5 \hr{hashtable.h}{src/interpreter/memory/hashtable.h}.
                    .5 \hr{array.c}{src/interpreter/memory/array.c}.
                    .5 \hr{array.h}{src/interpreter/memory/array.h}.
                    .5 \hr{stack.c}{src/interpreter/memory/stack.c}.
                    .5 \hr{stack.h}{src/interpreter/memory/stack.h}.
                .4 \hr{parser.c}{src/interpreter/parser.c}.
                .4 \hr{parser.h}{src/interpreter/parser.h}.
                .4 processing.
                    .5 \hr{implementations.c}{src/interpreter/processing/implementations.c}.
                    .5 \hr{implementations.h}{src/interpreter/processing/implementations.h}.
                .4 \hr{runtime.c}{src/interpreter/runtime.c}.
                .4 \hr{runtime.h}{src/interpreter/runtime.h}.
                .4 \hr{tokenizer.c}{src/interpreter/tokenizer.c}.
                .4 \hr{tokenizer.h}{src/interpreter/tokenizer.h}.
    }
\end{flushleft}}

@@

The file mapop.gperf is also a source file. It is used in conjunction with the GNU gperf
tool. The tool creates a perfect hash function, which is output to the file
src/interpreter/mappings/mapop.c. The perfect hash function is used to translate
single- or multi-character symbols, such as '!=' and '+', to their respective
interpreter internal representation in constant time.
[\cte{c_compilation}, \cte{hash}]

@@

The above figure is a representation of how different components of the entire
\name{} implementation interact and work together. Each next stage performs
operations on the output data of the proper. The src/interpreter/interpreter.c
file, here depicted as 'Interpreter', chains all stages together.

@@

\subsubsection{The Entry-Point}
The entry point of the project lies in the src/main.c file.
The file implements the C main function, which is the entry-point of any
C application in general. Since the file is so small, here is its source code:
\code{src/main.c}{1}{999}

For the readers who do not know how to read source code, C in particular, here is a short
explanation of the basic syntax of the C language based on the src/main.c file:
\begin{itemize}
    \item \emph{Preprocessor Instructions.} Lines beginning with hashtags are
        C preprocessor instructions, they are dealt with by the C preprocessor.
        Preprocessing is what the compiler does before actually compiling the
        code.

        The most common preprocessor instructions are the following.
        On one hand there is \expr{\#include <header.h>}
        to include various definitions from a header file, i.e. to paste
        its content at where the 'include' statement lies.
        On the other hand exists \expr{\#define A B} used to
        define A as an alias for B and replace every following occurrence of A
        with B.

    \item \emph{Functions.} Functions in C are defined the following way:
        \expr{<return type> <function name>([arguments]) { <function body> }}
        or with a concrete, working example:
        \expr{int myFunction(int *arg) \{ printf("arg: \%p, arg); \}.}
        In main.c there is one function. The function's name is 'main', its
        return type is 'int', it has two arguments 'argc' and 'argv' and a
        function body containing what the function does.

    \item \emph{Pointers.} Pointers are variables that store memory addresses
        of data. In C-like syntax they are defined with the star (*) symbol.
        An example from src/main.c is 'FILE *input;'. That means that input is not
        actually a FILE, but rather the memory address to where a FILE is.

        To dereference the pointer, i.e. to actually retrieve the value stored
        at the memory address it contains, another star could be used as such:
        'FILE value = *input;'.
\end{itemize}

For those who are further interested in the C programming language, there are
many internet articles or tutorials about reading and understanding or
writing and compiling C code. One such example is the article 'The C Beginner's
Handbook' provided by freeCodeCamp.org linked in the sources
on page \pageref{bibliography}.

Here is a very quick summary of what happens in the src/main.c file: the src/interpreter/interpreter.h
header file is included. In the main function, the program decides to either
read a program from the standard input or from a file. Then, the interpret function
from the said included header is called with the selected input.
[\cte{freeCodeCamp}]

\subsubsection{The Interpreter}
This section deals with the file src/interpreter/interpreter.c which guides
input data through the various stages of the \name{} interpreter in its
function 'interpret', which is called from the program's main function.

The 'interpret' function calls the parser to generate a syntax-tree from tokens retrieved
from the tokenizer implemented in src/interpreter/parser.c and 
src/interpreter/tokenizer.c respectively.

After that, variables with names are 'localized', meaning their names are
mapped to addresses in memory where their value is stored. Doing this before
bytecode compilation has a performance benefit. 
The mapping is done with a custom hashtable datastructure with average time
complexity of O(log n), n being the amount of different variables.

The syntax tree could now be interpreted directly, but as said, is instead converted
to an intermediate internal bytecode format by src/interpreter/bytecode.c.
This step is done so that code that is run multiple times does not convey
the problem of time overhead when traversing the syntax tree, but the current
implementation does not take advantage of this.

The final step is to process or execute said bytecode. This step is performed
by src/interpreter/runtime.c.

\subsubsection{The Tokenizer}
The tokenizer is implemented in src/interpreter/tokenizer.c. Its job is to
take source code in plain-text format as input and output so-called tokens.
These tokens represent groups of characters that belong together. In this paper's
implementation, each token has a strict type and a text content.
The different types of tokens and other things are defined in the header file
src/interpreter/tokenizer.h.

Here is an example of tokenization:
An input of 'std.test==23.4' would result in the following tokens according
to the tokenizer implementation of this paper:
\begin{enumerate}
    \item Token of type FIELD with content 'std.text'.
    \item Token of type SYMBOL with content '=='.
    \item Token of type NUMBER with content '23.4'.
\end{enumerate}
FIELD type tokens are later replaced by REFERENCE tokens during the stage of localization
and SYMBOL type tokens are mapped to EXPR by src/implementation/mappings/mapop.c
during parsing. NUMBER type tokens are replaced by an actual number type, which
is C's 'long double' type as specified in src/interpreter/bytecode.h.

@@

\subsubsection{The Parser}
The parser, implemented in src/interpreter/parser.c, generates a syntax-tree
from tokens. The algorithm used is inspired by the Pratt Parsing
algorithm first described by Vaughan Pratt in 1973, which is a kind of precedence
parser based on recursive descent. There is a great simple explanation of the algorithm
by Jonathan Apodaca on dev.io linked in the references on page 
\pageref{bibliography} which this paper's parser takes inspiration from.
[\cte{pratt}, \cte{devio}]

The syntax tree in the implementation of \name{} consists of 'stnode' type
pointers, as defined in src/interpreter/parser.h. Each such node depending
on its 'nodetype', declared in src/interpreter/tokenizer.h,
is either a value-node storing a pointer to some data, e.g. a number or text, or a
parent-node referencing two child nodes and an operator, e.g. two numbers as
children and a plus-operator to signify an addition.

The difficult part of implementing a parser is precedence. Precedence and
associativity define the order of execution of statements. A simple example
would be to perform multiplication before addition, e.g. '2+3*4' would
result in 14 and is not the same as '(2+3)*4'. Therefore, each operator is
assigned a precedence, or priority, as well as an associativity, which is
either 'Left to Right' or 'Right to Left'. An example of associativity is
\emph{'20 / 10 / 2'}. Both in C as well as in \name{}, the division operator is
left-to-right associative, meaning the stated expression is the same as
\emph{'(20 / 10) / 2'} and not as \emph{'20 / (10 / 2)'}. Both precedence 
and associativity are properties that are defined in the file mapop.gpref for
each operator.

The parser implementation in src/interpreter/parser.c may be difficult to
read and understand, which is why the most important functions are explained here:
\begin{itemize}
    \item \emph{advance.} This function reads a new token from the tokenizer.
    \item \emph{secondary.} The 'secondary' function deals with parsing single,
        simple values that make up the start of an expression, such as numbers,
        fields, strings and it handles brackets, such as indexing with square
        brackets, functions or code blocks with curly braces and sub-expressions
        in standard brackets.
    \item \emph{expr.} The 'expr' function calls 'secondary' which acts
        as the basis for a new expression. It then repeatedly checks
        if an operator follows the secondary token and creates a tree of
        EXPR type stnodes, each having children which are recursively parsed
        and information about the operator.
    \item \emph{parse.} The parse function is a small wrapper which only
        calls the 'expr' function, initiating the parsing process.
\end{itemize}

@@

\subsubsection{The Bytecode Compiler}
The bytecode compiler is part of the interpreter. Its job is to create a list
of byte-sized instructions and data from the syntax-tree the parser created.
Having such an intermediary bytecode format, instead of directly interpreting
the parse-tree can be beneficial. This is the case when the same piece of code
has to be run several times. Instead of generating a time and memory overhead 
having to walk along tree-nodes recursively, only the list of instructions in
the bytecode can be traversed, which is much faster.

@@

In src/interpreter/bytecode.c we can see, that first, the left operand and then
the right operand are added onto the stack of bytecode. Therefore, when
popping from the stack and evaluating instructions, the operands will have to
be read in reverse.
Here is an example:
\begin{enumerate}
    \item Assume the following mathematical expression in infix notation:
        \expr{\emph{'1 - 2 * 4 / (5 + 6)'}}
    \item In common prefix notation, also called Polish notation, the
        operators would now precede the operands as such:
        \expr{\emph{'- 1 (/ (* 2 4) (+ 5 6))'}}
    \item But in \name{} the operands are pushed onto the stack is reverse order.
        This would look like the following:
        \expr{\emph{'- (/ (+ 6 5) (* 4 2)) 1'}}
    \item The brackets are not needed, because for every operator, the amount
        of operands is known. This is the case as e.g. the subtraction operator
        and the negative operator, which are both a minus sign, are distinguished
        beforehand. This means the above statement would look like this:
        \expr{\emph{'- / + 6 5 * 4 2 1'}}
        This is an ideal abstraction of how data 
        is represented internally in \name{}.
\end{enumerate}

\subsubsection{The Runtime}
This is the last stage of interpreting program code as implemented in \name{}.
The runtime steps through the intermediary bytecode created by the bytecode
compiler and evaluates expressions by plugging in the provided values.
All available expressions are implemented in 
src/interpreter/processing/implementations.c.

The evaluation of expressions also works in a recursive manner. Here is an 
example of \name{}'s bytecode notation and how it would be evaluated at runtime:
\begin{itemize}
    \item Assume the input in \name{}'s bytecode to resemble the following expression: 
        \expr{\emph{'- 6 * 3 4'}}
        In infix notation, this would be equivalent to 
        \expr{\emph{'(4 * 3) - 6'}.}
    \item The runtime will read through the bytecode left-to-right. First it will
        notice the subtraction-operator and accordingly assume that two
        operands must follow.
    \item The processing function will call itsself
        and notice the 6, a number, and therefore a value. The function simply returns the 6
        to the subtraction expression. This 6 will be the right-hand operator
        of the subtraction.
    \item The subtraction will then call the processing function again to find
        the left operand. As the processing function reads a multiplication operator,
        it now recursively reads through that expression's right and then left
        operator, which would be 3 and for respectively and compute and return
        the result of the multiplication, which is 12.
    \item Now the subtraction knows its right- and left-hand operators, 6 and 12,
        and again, calculate and return the result, which is 6.
\end{itemize}

That is it. With all the expressions available, entire programs
can now be interpreted.

>>>>

% hyperref abbreviations
\newcommand{\hr}[2]{\hyperref[#2]{#1}}
\newcommand{\hrc}[1]{\hyperref[#1]{#1}}

@@

% Include C code from file #1 from lines #2 to #3
\newcommand{\sourcecode}[3] {
    \ver{
    \lstset{basicstyle=\scriptsize,keywordstyle=\color{blue},stepnumber=5,
        stringstyle=\color[rgb]{.3,.5,.1},commentstyle=\color{gray},
        morecomment=[l][\color{magenta}]{\#},
        postbreak=\mbox{\textcolor{red}{$\hookrightarrow$}\space}}
    \lstinputlisting[language=C,numbers=left,breaklines=true,
        frame=single,title=#1,firstnumber=#2,firstline=#2,lastline=#3]{#1}
    }
}

\newcommand{\code}[3] {
    \ver{

    \begin{minipage}{\linewidth}

    \lstset{basicstyle=\scriptsize,keywordstyle=\color{blue},stepnumber=5,
        stringstyle=\color[rgb]{.3,.5,.1},commentstyle=\color{gray},
        morecomment=[l][\color{magenta}]{\#},
        postbreak=\mbox{\textcolor{red}{$\hookrightarrow$}\space}}
    \lstinputlisting[language=C,numbers=left,breaklines=true,
        frame=single,title=\hrc{#1},
        firstnumber=#2,firstline=#2,lastline=#3]{#1}

    \end{minipage}
    }
}

@@

One more goal is to learn and understand the C programming language, in which
the interpreter of \name{} will be written in.

\section{Theory}
\subsection{Different Types of Interpreters}
In computing, there are several types of interpreters that behave differently.
Here are some examples and their use-cases:
\begin{itemize}
    \item \emph{Just in time Compiler.} JIT-compilers blur the line between
        classical interpreters and compilers. As the name suggests, code is
        being compiled at run time of the program. The JIT-compiler can optimize
        compilation of more-often used code and therefore adapt itself to
        behave somewhat close to optimally for a given use-case. Today, this 
        technique is often used instead of classical interpreters, as
        JIT-compilers come with only advantages. The most popular example of
        a JIT-compiler would be Google's V8 JavaScript engine used
        in NodeJS and any Chromium browsers.
    \item \emph{Bytecode Interpreter.} Such interpret the bytecode that was
        output by a bytecode compiler from a given piece of code. The bytecode
        is an intermediate representation of the program, used to speed up
        both compilation and interpretation. Such an intermediary bytecode is
        useful for ensuring platform independence or portability, as it will be
        the same across any machine, the only difference being the interpreter.
        The most popular example would be the Java Virtual Machine,
        a bytecode interpreter for java bytecode, also used by many other languages
        such as Kotlin, Scala or Groovy.
    \item \emph{Abstract Syntax Tree interpreter.} Source code can be
        transformed into an AST or parse tree by a parser. A simple example
        can be found in section \ref{simple_interpreter}
        on page \pageref{simple_interpreter}. Compared to bytecode interpreters
        there is a large time overhead when performing syntax related computation
        or visiting tree nodes recursively.
    \item \emph{Self interpreter.} These are interpreters that interpret the
        language they themselves are written in. This requires a program
        written in another language running an interpreter for the wanted
        language, in which runs another interpreted for and written in said
        language. Using a host language to initiate such system is called
        bootstrapping.

@@

The problem to solve in this paper is concisely to make a computer understand
human-written and -readable text, called source code, and follow the instructions
it proposes.
Most interpreters follow a plain scheme to go about this problem, but the
actual implementation of each of the steps can be heavily customized.
Here is the problem, broken down into three steps as commonly done in other
popular programming languages:
\begin{enumerate}\pagelabel{simple_interpreter}
    \item To create meaningful groups of characters from the input string. This
        process is called 'tokenization' and is done with the component called
        tokenizer or lexer. As an example, the string 'aa + bb' could result
        in three tokens, namely 'aa', '+' and 'bb'. In this case, space characters
        are ignored. 
        This is only an example and the grouping of characters could be done 
        in any thinkable, logical way.
    \item To order the tokens into parse-tree, also called syntax-tree, as
        specified by rules. These rules are called a grammar.
        Creating an order is important, as some calculations depend on others, e.g. multiplication
        is always done before addition. An IT data structure that can accommodate
        such requirements is the binary tree. A binary tree consists of tree-nodes
        that each have a maximum of two other tree nodes as their children, and
        a tree node whose child they are, the parent node.
    \item To interpret the parse tree, traversing through it and following the
        instructions each of the tree-nodes represent. The outcome
        will be the result of the computation the program describes.
\end{enumerate}

@@

Here is a file tree overview of the project which will be referenced
several times in the following sections of this paper. In the depicted file tree
only source files are shown. No build files, executable files or files that
are of no further importance regarding \name are depicted.
\ver{\begin{flushleft}
    \scriptsize\setlength{\DTbaselineskip}{7pt}
    \dirtree{%
        .1 project root.
        .2 \hr{.gitignore}{.gitignore}.
        .2 \hr{README.md}{README.md}.
        .2 \hr{Makefile}{Makefile}.
        .2 \hr{mapop.gperf}{mapop.gperf}.
        .2 pseudocode.
            .3 \hr{parser.c}{pseudocode/parser.c}.
        .2 examples.
            .3 \hr{mandelbrot.nts}{examples/mandelbrot.nts}.
            .3 \hr{fibonacci.nts}{examples/fibonacci.nts}.
            .3 \hr{exec.nts}{examples/exec.nts}.
        .2 src.
            .3 \hr{main.c}{src/main.c}.
            .3 interpreter.
                .4 \hr{bytecode.c}{src/interpreter/bytecode.c}.
                .4 \hr{bytecode.h}{src/interpreter/bytecode.h}.
                .4 error.
                    .5 \hr{error.c}{src/interpreter/error/error.c}.
                    .5 \hr{error.h}{src/interpreter/error/error.h}.
                .4 \hr{interpreter.c}{src/interpreter/interpreter.c}.
                .4 \hr{interpreter.h}{src/interpreter/interpreter.h}.
                .4 libraries.
                    .5 \hr{libraries.c}{src/interpreter/libraries/libraries.c}.
                    .5 \hr{libraries.h}{src/interpreter/libraries/libraries.h}.
                    .5 stdlib.
                        .6 \hr{stdlib.c}{src/interpreter/libraries/stdlib/stdlib.c}.
                        .6 \hr{stdlib.h}{src/interpreter/libraries/stdlib/stdlib.h}.
                .4 \hr{localizer.c}{src/interpreter/localizer.c}.
                .4 \hr{localizer.h}{src/interpreter/localizer.h}.
                .4 mappings.
                    .5 \hr{operations.c}{src/interpreter/mappings/operations.c}.
                    .5 \hr{operations.h}{src/interpreter/mappings/operations.h}.
                    .5 \hr{mapop.c}{src/interpreter/mappings/mapop.c}.
                    .5 \hr{mapop.h}{src/interpreter/mappings/mapop.h}.
                .4 memory.
                    .5 \hr{hashtable.c}{src/interpreter/memory/hashtable.c}.
                    .5 \hr{hashtable.h}{src/interpreter/memory/hashtable.h}.
                    .5 \hr{array.c}{src/interpreter/memory/array.c}.
                    .5 \hr{array.h}{src/interpreter/memory/array.h}.
                    .5 \hr{stack.c}{src/interpreter/memory/stack.c}.
                    .5 \hr{stack.h}{src/interpreter/memory/stack.h}.
                .4 \hr{parser.c}{src/interpreter/parser.c}.
                .4 \hr{parser.h}{src/interpreter/parser.h}.
                .4 processing.
                    .5 \hr{implementations.c}{src/interpreter/processing/implementations.c}.
                    .5 \hr{implementations.h}{src/interpreter/processing/implementations.h}.
                .4 \hr{runtime.c}{src/interpreter/runtime.c}.
                .4 \hr{runtime.h}{src/interpreter/runtime.h}.
                .4 \hr{tokenizer.c}{src/interpreter/tokenizer.c}.
                .4 \hr{tokenizer.h}{src/interpreter/tokenizer.h}.
    }
\end{flushleft}}

@@

% TODO: say sth about README.md and expand it

The file mapop.gperf is also a source file. It is used in conjunction with the GNU gperf
tool. The tool creates a perfect hash function, which is output to the file
src/interpreter/mappings/mapop.c. The perfect hash function is used to translate
single- or multi-character symbols, such as '!=' and '+', to their respective
interpreter internal representation in constant time.
[\cte{c_compilation}, \cte{gperf}]

@@

The above figure is a representation of how different components of the entire
\name{} implementation interact and work together. Each next stage performs
operations on the output data of the proper. The \hrc{src/interpreter/interpreter.c}
file, here depicted as 'Interpreter', chains all stages together.

@@

\subsubsection{The Entry-Point}
The entry point of the project lies in the \hrc{src/main.c} file.
The file implements the C main function, which is the entry-point of any
C application. Since the file is so small, here is its source code:
\code{src/main.c}{1}{999}

For the readers who do not know how to read source code, C in particular, here is a short
explanation of the basic syntax of the C language taking the \hrc{src/main.c} file
as an example:
\begin{itemize}
    \item \emph{Preprocessor Instructions.} Lines beginning with hashtags are
        C preprocessor instructions, they are dealt with by the C preprocessor.
        Preprocessing is what the compiler does before actually compiling the
        code.

        The most common preprocessor instructions are the following.
        On one hand there is \expr{\#include $<$header.h$>$}
        to include various definitions from a header file, i.e. to paste
        its content at where the 'include' statement lies.
        On the other hand exists \expr{\#define A B} used to
        define A as an alias for B and replace every following occurrence of A
        with B.

    \item \emph{Functions.} Every C function has a return type, a
        function name, arguments and a function body in that order.
        Here is an example of a C function:
        \code{pseudocode/function.c}{1}{999}
        This would be an example of a main function, similar to the one
        in \hrc{src/main.c}. It returns a value
        of type 'int', has two arguments, 'argc' and 'argv', and a function body
        which, in the case of the example above, prints the value of the 'argc' argument.

    \item \emph{Pointers.} Pointers are variables that store memory addresses
        of data. In C-like syntax they are defined with the star '*' symbol.
        An example from \hrc{src/main.c} is 'FILE *input;'. That means that input is not
        actually a FILE, but rather a memory address pointing to a spot in memory where a FILE is.
        To dereference the pointer, i.e. to copy the value stored
        at the memory address into a variable on the stack, another star is used like this:
        'FILE value = *input;'.
\end{itemize}

For those who are further interested in the C programming language, there are
countless internet articles or tutorials about reading and understanding or
writing and compiling C code. One such example is the article 'The C Beginner's
Handbook' provided by freeCodeCamp.org linked in the sources
on page \pageref{bibliography}.

Following is a very quick summary of the few 
things happening in the \hrc{src/main.c} file.
First, the \hrc{src/interpreter/interpreter.h} header file is included. 
The main function decides whether to
read a program from the standard input or from a file. 
Then, the interpret function
from the said included header is called with the selected input.
[\cte{freeCodeCamp}]

\subsubsection{The Interpreter}
This section deals with the file 
\hrc{src/interpreter/interpreter.c} which guides
input data through the various stages of the \name{} interpreter as 
seen in figure \ref{f_diagram} on page \pageref{f_diagram}.
Following is a breakdown of the 'interpret' function from said file.

The 'interpret' function calls the parser to generate a syntax-tree from tokens retrieved
from the tokenizer implemented in \hrc{src/interpreter/parser.c} and 
\hrc{src/interpreter/tokenizer.c} respectively.

After that, variables with names are 'localized', meaning their names are
mapped to addresses in memory where their value is stored. Doing this before
bytecode compilation has a performance benefit. 
The mapping is done with a custom hashtable datastructure with average time
complexity of O(log n), n being the amount of different variables.
The hashtable implementation used in \hrc{src/interpreter/memory/hashtable.c}
uses binary search to locate elements variable-sized arrays in the hashtable's buckets.
For collision detection, i.e. to circumvent problems arising when two keys have the
same hash, separate chaining is used. In this approach, linked lists with
key-value pairs are used for each array index and keys are compared directly
to find exact matches.
[\cte{separate_chaining}]

The syntax tree could now be interpreted directly, but as said, is instead converted
to an intermediate internal bytecode format by \hrc{src/interpreter/bytecode.c}.
This step is done so that code that is run multiple times does not convey
the problem of time overhead when traversing the syntax tree, but the current
implementation does not take advantage of this.

The final step is to process or execute said bytecode. This step is performed
by \hrc{src/interpreter/runtime.c}.

\subsubsection{The Tokenizer}
The tokenizer is implemented in \hrc{src/interpreter/tokenizer.c}. Its job is to
take source code in plain-text format as input and output so-called tokens.
These tokens represent groups of characters that belong together. In this paper's
implementation, each token has a strict type and a text content.
The different types of tokens and other things are defined in the header file
\hrc{src/interpreter/tokenizer.h}.

Here is an example of tokenization:
An input of 'std.test==23.4' would result in the following tokens according
to the tokenizer implementation of this paper:
\begin{enumerate}
    \item Token of type FIELD with content 'std.text'.
    \item Token of type SYMBOL with content '=='.
    \item Token of type NUMBER with content '23.4'.
\end{enumerate}
FIELD type tokens are later replaced by REFERENCE tokens, effectively mapping
variable names to addresses in memory.
SYMBOL type tokens are mapped to operators such as plus or minux by \hrc{src/implementation/mappings/mapop.c}
during parsing. NUMBER type tokens are replaced by an actual number type, which
is C's 'long double' type as specified in \hrc{src/interpreter/bytecode.h}.

@@

\subsubsection{The Parser}
The parser, implemented in \hrc{src/interpreter/parser.c}, generates a syntax-tree
from tokens. The algorithm used is inspired by the Pratt Parsing
algorithm first described by Vaughan Pratt in 1973, which is a kind of precedence
parser based on recursive descent. There is an explanation of the algorithm
written by Jonathan Apodaca on dev.io linked in the references on page 
\pageref{bibliography} which this paper's parser takes inspiration from.
[\cte{pratt}, \cte{devio}]

The syntax tree in the implementation of \name{} consists of 'stnode' type
pointers defined in \hrc{src/interpreter/parser.h}. Each such node depending
on its 'nodetype', declared in \hrc{src/interpreter/tokenizer.h},
is either a value-node storing a pointer to data, e.g. a number or text, or a
parent-node referencing two child nodes and an operator, e.g. two numbers as
children and a plus-operator to signify an addition.

The difficult part of implementing a parser is precedence. Precedence together
with associativity define the order of execution of statements. A simple example
would be to perform multiplication before addition, e.g. \emph{2+3*4} would
result in \emph{14} and is not the same as \emph{(2+3)*4}.
Assosiativity is either 'Left to Right' or 'Right to Left'. An example of associativity is
\emph{20/10/2}. Both in C as well as in \name{}, the division operator is
left-to-right associative, meaning the stated expression is the same as
\emph{(20/10)/2} and not as \emph{20/(10/2)}. Both precedence 
and associativity are properties that are defined in the file \hrc{mapop.gpref} for
each operator.

The parser implementation in \hrc{src/interpreter/parser.c} may be difficult to
read and understand, which is why the most important functions are explained here:
\begin{itemize}
    \item \emph{advance.} This function reads a new token from the tokenizer.
    \item \emph{secondary.} The 'secondary' function deals with parsing single,
        simple values that make up the start of an expression, such as numbers,
        fields, strings and it handles brackets, such as indexing with square
        brackets, functions or code blocks with curly braces and sub-expressions
        in standard brackets.
    \item \emph{expr.} The 'expr' function calls 'secondary' which acts
        as the basis for a new expression. It then repeatedly checks
        if an operator follows the secondary token and creates a tree of
        EXPR type stnodes, each having children which are recursively parsed
        and information about the operator.
    \item \emph{parse.} The parse function is a small wrapper which only
        calls the 'expr' function, initiating the parsing process.
\end{itemize}

@@

\subsubsection{The Bytecode Compiler}
The bytecode compiler is an internal part of the interpreter, located in the file
\hrc{src/interpreter/bytecode.c}. Its job is to create a list
of byte-sized instructions and data from the syntax-tree the parser created.
Having such an intermediary bytecode format, instead of directly interpreting
the parse-tree can be beneficial. This is the case when the same piece of code
has to be run several times. Instead of generating a time and memory overhead 
having to walk along tree-nodes recursively, only the list of instructions in
the bytecode can be traversed, which is much faster.

@@

In \hrc{src/interpreter/bytecode.c} we can see that first the left operand and then
the right operand are added onto the stack of bytecode. Therefore, when
popping from the stack and evaluating instructions, the operands will have to
be read in reverse.
Here is an example:
\begin{enumerate}
    \item Assume the following mathematical expression in infix notation:
        \expr{\emph{'1 - 2 * 4 / (5 + 6)'}}
    \item In common prefix notation, also called Polish notation, the
        operators would now precede the operands in this manner:
        \expr{\emph{'- 1 (/ (* 2 4) (+ 5 6))'}}
    \item But in \name{} the operands are pushed onto the stack is reverse order.
        This would look like the following:
        \expr{\emph{'- (/ (+ 6 5) (* 4 2)) 1'}}
    \item The brackets are not needed, because for every operator, the amount
        of operands is known. This is the case as e.g. the subtraction operator
        and the negative operator, which are both a minus sign, are distinguished
        beforehand. This means the above statement would look like this:
        \expr{\emph{'- / + 6 5 * 4 2 1'}}
        This is an ideal abstraction of how data 
        is represented internally in \name{}.
\end{enumerate}
[\cte{polish_notation}]

\subsubsection{The Runtime}
This is the last stage of interpreting program code as implemented in \name{}.
The runtime steps through the intermediary bytecode created by the bytecode
compiler and evaluates expressions by plugging in the provided values.
All available expressions are implemented in 
\hrc{src/interpreter/processing/implementations.c}.

The evaluation of expressions also works in a recursive manner. Here is an 
example of \name{}'s bytecode notation and how it would be evaluated at runtime:
\begin{itemize}
    \item Assume the input in \name{}'s bytecode to resemble the following expression: 
        \expr{\emph{'- 6 * 3 4'}}
        In infix notation, this would be equivalent to 
        \expr{\emph{'(4 * 3) - 6'}.}
    \item The runtime will read through the bytecode left-to-right. First it will
        notice the subtraction-operator and accordingly assume that two
        operands must follow.
    \item The processing function will call itself
        and notice the 6, a number, and therefore a value. The function simply returns the six
        to the subtraction expression. This six will be the right-hand operator
        of the subtraction.
    \item The subtraction will then call the processing function again to find
        the left operand. As the processing function reads a multiplication operator,
        it now recursively reads through that expression's right and then left
        operator, which would be 3 and for respectively and compute and return
        the result of the multiplication, which is 12.
    \item Now the subtraction knows its right- and left-hand operators, 6 and 12,
        and again, calculate and return the result, which is 6.
\end{itemize}

That is it. With all expressions that are currently implemented, entire programs
can be interpreted. Examples of such programs can be found in the \emph{examples} directory
of the source code.
